---
title: 'Homework 3'
author: 'Shun-Lung Chang, Dilip Hiremath'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, comment = '', fig.align = 'center', fig.height = 4, fig.width = 4)
```

```{r}
load('data/bankmarketing.Rdata')
```

## 1. Create a logistic regression model to predict the efficiency of the marketing campaign using y as dependent variable and no predictor, just the intercept.

```{r}
mod_1 <- glm(y ~ 1, data = bankmarketing, family = binomial(link = 'logit'))
summary(mod_1)
```

### (a) (half a point) How large is the AIC score for this model?

The AIC score is 3233.

```{r}
AIC(mod_1)
```

### (b) (half a point) How large is the BIC score for this model?

The BIC score is 3239.417.

```{r}
BIC(mod_1)
```
### (c) (1 point) Compute the log odds for the mean response and compare it to the coefficient estimate for the intercept in this model [hint: Be careful about the internal numeric coding of the variable y in the data set].

The log odds is computed as follows and the value is -2.0383, which is the same as the intercept in the model.

```{r}
log(sum(bankmarketing$y == 'yes') / sum(bankmarketing$y == 'no'))
```
## 2. Add duration as a predictor to the model.

```{r}
mod_2 <- glm(y ~ duration, data = bankmarketing, family = binomial(link = 'logit'))
summary(mod_2)
```

### (a) (half a point) In comparison to the naive model, by how much has the AIC changed?

The AIC score drops by 527.2476 after *duration* is added to the model.

```{r}
AIC(mod_2) - AIC(mod_1)
```

### (b) (half a point) Is last contact duration (duration) a significant predictor in this model for modeling the probability of subscribing to a term deposit?

The small p-value implys that the variable *duration* is significant. 

### (c) (half a point) With growing duration of the last contact are clients less likely or are they more likely to subscribe to a term deposit?

The positive coefficient suggests that customer with larger duration will be more likely to have a term deposit.

### (d) (half a point) According to the second model, what is the estimated probability of subscribing to a term deposit for a client who immediately terminated the last contact (i.e. duration equals 0)?

The probability is 0.037 if a client with 0 duration.

```{r}
predict(mod_2, data.frame(duration = 0), type = 'response')
```

## 3. Using the second model with duration as predictor,

### (a) (1 point) compute the halfway point, i.e the last contact duration at which the estimated probability of subscribing to a term deposit equals 0.5.

Based on our model, which is $ln(\frac{p}{1-p}) = \hat{\beta_o} + \hat{\beta_1} * duration$, if the estimated probability is 0, then we have $\hat{\beta_o} + \hat{\beta_1} * duration = 0$. Therefore, the duration at the halfway point is $duration = -\frac{\hat{\beta_o}}{\hat{\beta_1}}$, whose value in this model is 917.2802.

```{r}
d = unname(-mod_2$coefficients[1] / mod_2$coefficients[2])
d
```

We can as well predict the estimated probability when duration is 917.2802. The probability as expected is 0.5. 

```{r}
predict(mod_2, data.frame(duration = d), type = 'response')
```

### (b) (1 point) Compute the slope of the tangent to the regression curve at the halfway point.

A straight forward way to compute the slope of a function is to take its derivative. Also, the derivative of the model, $P(y = 1) = \frac{1}{1 + e^{-(\hat{\beta_o} + \hat{\beta_1} * X)}}$, is $\beta_1 * \frac{e^{-(\hat{\beta_o} + \hat{\beta_1} * X)}}{(1 + e^{-(\hat{\beta_o} + \hat{\beta_1} * X)})^2} = \beta_1 * P(y = 1) * (1 - P(y = 1))$. Accoring to the formula, we know the slope at the halfway point, namely $P(y = 1) = 0.5$, is 0.0008873882.

```{r}
unname(mod_2$coefficients[2]) * 0.5 * (1 - 0.5)
```

## 4. Compute a logistic regression model for subscribing to a term deposit using age, marital and duration as predictors.

```{r}
mod_3 <- glm(y ~ age + marital + duration, data = bankmarketing, 
             family = binomial(link = 'logit'))
summary(mod_3)
```

### (a) (1 point) Calculate the predictive probability of subscribing to a term deposit for a married client at the mean values of the numeric predictors.

Given the conditions, the predictive probability is 0.0734.

```{r}
predict(mod_3, data.frame(age = mean(bankmarketing$age),
                          marital = 'married',
                          duration = mean(bankmarketing$duration)),
        type = 'response')
```

### (b) In the above model, calculate the effect on the probability of subscribing when keeping all other predictors constant and

#### i. (half a point) changing duration from the mean score to 300 seconds;

The estimated probability is increased to 0.0825 as the duration is 300.

```{r}
predict(mod_3, data.frame(age = mean(bankmarketing$age),
                          marital = 'married',
                          duration = 300),
        type = 'response')
```

#### ii. (half a point) changing age from the mean score to one standard deviation above the mean score.

The estimated probability is increased to 0.0911 if the age is one standard deviation above the mean value.

```{r}
predict(mod_3, data.frame(age = mean(bankmarketing$age) +
                              sd(bankmarketing$age),
                          marital = 'married',
                          duration = mean(bankmarketing$duration)),
        type = 'response')
```

## 5. (2 points) Compute a logistic regression model for the decision to subscribe to a term deposit using duration, campaign, and the interaction between the two. How do you interpret the regression coefficients? Are these interpretations meaningful? Give reasons for your answer!

The following table indicates that when duration is increased by one unit, the change of odds for the event y is 'yes' is $exp(0.00277087 + 0.00032618 \times campaign)$. On the other hand, if number of campaign is increased by one, the change of odds is $exp(-0.27602084 + 0.00032618 \times duration)$. Because of the small p-value, we can say that there exists interactions between the two variables.

```{r}
mod_4 <- glm(y ~ duration + campaign + duration * campaign, 
             data = bankmarketing,
             family = binomial(link = 'logit'))
summary(mod_4)
```

## 6. (2 points) Center the variables duration and campaign and re-build the model built in question 5. How do you interpret the regression coefficients? Are these interpretations meaningful? Give reasons for your answer! Draw the effects plot for this model and interpret!

As the table below shows, the coefficient of the interaction term remains the same although the variables are centered. 

```{r}
c_duration <- scale(bankmarketing$duration, scale = FALSE)
c_campaign <- scale(bankmarketing$campaign, scale = FALSE)

mod_5 <- glm(bankmarketing$y ~ c_duration + c_campaign + c_duration * c_campaign,
            family = binomial(link = 'logit'))
summary(mod_5)
```

```{r}
library(sjPlot)
```

## 7. (2 points) Create a logistic regression model using y as dependent variable and all available predictors, except day and month. Which predictors are significant? Do the estimated coefficients make common sense to you?

Under significance level of 5%, the significant variables are *housing*, *loan*, *contact (if unknown)*, *duration*, *campaign*, and *poutcome (if success)*.

The coefficients of these variables seem valid. For the variables with a positive sign-duration, poutcome (if success)-, if clients' last contract was longer, it is more possible for them to have a term deposit since they might be satisfied with contract. We can also understand that a positive previous campaign outcome will lead to higher chance for a client to make a term deposit.

On the other hand, if people have a housing or personal load, it is anticipatable that they would not make a term deposit. Also, if the contact communication type is unknown, we may assume that the campaign was not made, and hence the client did not subscribe a term deposit. Interestingly, a higher number of campaign contacts will prevent people from subscribing a term deposit in that people might feel irritated if they had frequent contacts with campaign representatives.

```{r}
mod_6 <- glm(y ~ . - day - month, data = bankmarketing,
             family = binomial(link = 'logit'))
summary(mod_6)
```

## 8. (2 points) Starting with the model of Question 7 use the automatic backward/forward selection method to derive a suitable model. Report the significant predictors and the AIC score of the resulting model.

If we consider significance level of 5%, the significant predictors of the final model are *job (if retired)*, *housing*, *loan*, *contact (if unknown)*, *duration*, *campaign*, and *poutcome (if success)*. Additionally, the model's AIC score is 2330.4.

```{r, results=FALSE}
mod_7 <- step(mod_6, direction = 'both')
```
```{r}
summary(mod_7)
```

## 9. (2 points) Draw a box plot of the residuals and look for extreme outliers. Remove the outlier and re-run the model you have obtained in Question 8. Which changes in the model are to be noted?

The boxplots shows the distribution of the residuals of the model obtained in Task 8. In this task, we define outliers as the residuals with a absolute value greater than 2.8. By doing so, a new model could be obtained after we removed the outliers. The AIC score was reduced from 2330.4 to 2277.3; the model was improved by removing the outliers.

```{r}
res <- residuals(mod_7)
boxplot(res, horizontal = TRUE, 
        main = 'Boxplot of Residuals',
        xlab = 'residual')
```

```{r}
bankmarketing_2 <- subset(bankmarketing, !abs(res) > 2.8)
mod_8 <- glm(formula = y ~ job + marital + education + 
                 housing + loan + contact + duration + campaign +
                 poutcome, 
             family = binomial(link = "logit"), 
             data = bankmarketing_2)
summary(mod_8)
```
