---
title: "Homework 4"
author: "Shun-Lung Chang, Dilip Hiremath"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, comment = '', fig.align = 'center', fig.height = 4, fig.width = 4)
```

```{r}
library(foreign)
hsb <- read.dta('data/hsbdemo.dta')
```

## 1. Cross-tabulate the variables ses and prog.

```{r}
ct <- table(hsb$prog, hsb$ses) / nrow(hsb)
ct
```
### (a) (half a point) Which program was chosen by the largest fraction of students with high socio-economic status?

### (b) (half a point) How many percent of students with low socio-economic status selected the general program?



### (c) (half a point) In the academic program are there more students with middle socioeconomic status than students with high socio-economic status?

### (d) (half a point) What is the least-frequent combination of the two variables?

## 2. You continue with your analysis of the relationship between ses and prog.

### (a) (half a point) Draw a mosaicplot visualising the contingency table of program choice and socio-economic status.

```{r}
mosaicplot(ct, main = 'Program vs. Socio-economic Status')
```

### (b) (1.5 points) Are students with low ses less likely (as measured in odds) to choose the academic program than students with higher socio-economic status? Calculate the odds ratios for choosing the academic program comparing students with low ses to students with middle ses and to students with high ses. [hint: use the command loddsratio in the package vcd. First, aggregate the variable prog into a binary variable indicating whether the student has chosen an academic program yes or no.]

```{r}
library(vcd)
hsb$academic_yes <- hsb$prog == 'academic'

ct_2 <- table(hsb$academic_yes, hsb$ses) / nrow(hsb)
loddsratio(ct_2)
```

```{r, echo=FALSE}
hsb$academic_yes <- NULL
```

## 3. Now, you assess the relationship between prog and ses using the $\chi^2$-statistic.

### (a) (1 point) Calculate the $\chi^2$-test to assess the relationship between ses and prog. Is the relationship statistically significant?

```{r}
chisq.test(hsb$prog, hsb$ses)
```

### (b) (1 point) Calculate the expected frequencies under the assumption that socio-economic status has no effect on program choice. For which cells are expected frequencies higher than the observed ones?

```{r}
expected <- chisq.test(hsb$prog, hsb$ses)$expected
expected - (ct * nrow(hsb))
```

## 4. In the following, perform the last analysis separately for female and male students.

```{r}
hsb_m <- hsb[hsb$female == 'male', ]
hsb_f <- hsb[hsb$female == 'female', ]
```

### (a) (half a point) Calculate the $\chi^2$-test to assess the relationship between ses and prog.

```{r}
chisq.test(hsb_m$prog, hsb_m$ses)
```

```{r}
chisq.test(hsb_f$prog, hsb_f$ses)
```

### (b) (half a point) Calculate the expected frequencies under the assumption that socio-econmic status has no effect on program choice. For which cells are expected frequencies higher than the observed ones?

```{r}
# male
observed_m <- table(hsb_m$prog, hsb_m$ses)
chisq.test(hsb_m$prog, hsb_m$ses)$expected - observed_m
```

```{r}
# female
observed_f <- table(hsb_f$prog, hsb_f$ses)
chisq.test(hsb_f$prog, hsb_f$ses)$expected - observed_f
```

### (c) (half a point) Do the results differ for the two sexes?

### (d) (half a point) Visualise the relationships using mosaicplots. Get any differences between females and males in relation to socio-economic status and program choice visible in the plots?

```{r, fig.width=5}
mosaicplot(observed_m, 'Program vs. Socio-economic Status (Male)')
```

```{r, fig.width=5}
mosaicplot(observed_f, 'Program vs. Socio-economic Status (Female)')
```

## 5. Create a multinomial logistic regression model using prog as dependent variable and the following predictors: female, ses, schtype, read, write, math, science, honors, awards. [hint: use the function multinom in hte package nnet.]

```{r}
library(nnet)
mod_1 <- multinom(prog ~ . - id - socst - cid, 
                  data = hsb)
summary(mod_1)
```

### (a) (half a point) How large is the AIC score for this model?

The AIC score is 358.8871.

```{r}
AIC(mod_1)
```

### (b) (1.5 points) The default output does not include p-values. Compute p-values based on the Wald-test statistics and determine the coefficients that are statistically significantly different from zero!

As the table below shows, the variables *academic:(Intercept)*, *academic:math*, *academic:science* and *vocation:sesmiddle* are significant under significance level of 5%.

```{r}
# compute p-values by definition
# z <- summary(mod_1, Wald.ratios = TRUE)$Wald.ratios
# p <- (1 - pnorm(abs(z))) * 2
# p

# compute p-values by package function
library(AER)
coeftest(mod_1)[, 4] < 0.05
```

## 6. Using the model from the previous question and the backward strategy with criterion AIC for variable selection, determine the significant coefficients in the resulting model.

```{r, results=FALSE}
mod_2 <- step(mod_1, direction = 'backward')
```
```{r}
summary(mod_2)
```

### (a) (1 point) Which predictors are included in the resulting model? 

According to the table above, the variables *ses*, *schtyp*, *read*, *math* and *science* are included in the final model.

### (b) (half a point) What is the BIC score of the resulting model?

The BIC score is 397.1684.

```{r}
BIC(mod_2)
```
### (c) (half a point) What is the log-likelihood score of this model?

The log-likelihood score is -161.496.

```{r}
logLik(mod_2)
```

## 7. (2 points) Using the final model that resulted in Question 6 predict the probabilities for the three program types for the combination of all factor levels and the average score of numeric predictors in the model.

The table below indicates academic is the most possible outcome in all combinations. But the students in the private school are more likely to take an academic program than that in the public school.

```{r}
d <- expand.grid(ses = c('low', 'middle', 'high'),
                schtyp = c('private', 'public'),
                read = mean(hsb$read),
                math = mean(hsb$math),
                science = mean(hsb$science))
```
```{r}
pred <- predict(mod_2, newdata = d, type = 'probs', se = TRUE)
cbind(d, pred)
```

## 8. (2 points) Again using the final model that resulted in Question 6, we now want to investigate the specific dependency on the math score. Generate new data such that you have for each combination of factor levels a total of 51 math scores running from 30 to 80 in increments of one. The other numeric predictors enter again with their mean score into the prediction. Compute the predictions and average them for each level of socio-economic status.

The table below shows the average probabilities of taking different programs with respect to different socio-economic levels. As can be seen from the table, taking an academic program is the most probable in three socio-economic levels. Furthermore, people with a high socio-economic status is more likely to have an academic program.

```{r, fig.width=4.5}
d <- expand.grid(ses = c('low', 'middle', 'high'),
                schtyp = c('private', 'public'),
                read = mean(hsb$read),
                math = 30:80,
                science = mean(hsb$science))

pred <-  predict(mod_2, newdata = d, type = 'probs', se = TRUE)

bind_d <- cbind(d, pred)
aggregate(bind_d[, 6:8], by = list(bind_d$ses), FUN = 'mean')
```
